---
title: "Practical Machine Learning --- Final Project Report"
author: "José Díaz"
date: "Saturday, March 21, 2015"
output: html_document
---

## Background

The following lines comprise the final project report for the Practical Machine Learning class from the Johns Hopkins University and Coursera. 

## Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the main goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset). 

## Data Source

The training data for this project are available here: 
<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv> 
The test data are available here: 
<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

## Purpose

The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. It is possible to use any of the other variables to predict with. At the end, a report describing how the model was built is presented, how you the cross validation was employed, what he expected out of sample error is, and why the choices made were done. Finally,  the prediction model which was buil will be used to predict 20 different test cases. 

* The submission will consist of a link to a Github repo with the R markdown and compiled HTML file describing the analysis. ( < 2000 words and the number of figures to be less than 5)*
*Apply the machine learning algorithm to the 20 test cases available in the test data above.* 

## Reproducibility 

Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis. 

# Initial Steps

For the sake of assuring the reproducibility in the results obtained, the following packages must be loaded into the working environment and the same pseudo random seed. 


```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```

Finally, load the same seed with the following line of code:

```{r}
set.seed(48993)
```

# Retreiving The Data

The training data set can be found on the following URL:

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```

The testing data set can be found on the following URL:

```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

Loading the data into the memory

```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

# Splitting the Training Data Set into two

Partioning Training data set into two data sets, following a general principle: 60% for myTraining, 40% for myTesting:


```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]
dim(myTraining)
dim(myTesting)
```

## Cleaning the Data

The following transformations were used to clean the data:

### Transformation 1: Cleaning NearZeroVariance Variables Run this code to view possible NZV Variables:


```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
```

The following code was used to create another subset without NZV variables:

```{r}
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt", "kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", 
"skewness_roll_dumbbell", "skewness_pitch_dumbbell", "skewness_yaw_dumbbell", 
"max_yaw_dumbbell", "min_yaw_dumbbell", "amplitude_yaw_dumbbell", "kurtosis_roll_forearm",
"kurtosis_picth_forearm", "kurtosis_yaw_forearm", "skewness_roll_forearm", 
"skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm", "max_yaw_forearm",
"min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm", "amplitude_yaw_forearm",
"avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm", "avg_pitch_forearm", 
"stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm", "stddev_yaw_forearm",
"var_yaw_forearm")

myTraining <- myTraining[!myNZVvars]
dim(myTraining)
```

### Transformation 2: Killing first column of Dataset - ID Removing first ID variable so that it does not interfer with ML Algorithms:

```{r}
myTraining <- myTraining[c(-1)]
```

### Transformation 3: Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA's I'm going to leave them out:

```{r}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
                trainingV3 <- trainingV3[ , -j] #Remove that column
            }   
        } 
    }
}

dim(trainingV3)
```

Going  back to the previous set:

```{r}
myTraining <- trainingV3
rm(trainingV3)
```

Now let us do the exact same 3 transformations but for our myTesting and testing data sets.

```{r}

clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2,]
dim(myTesting)
dim(testing)
```

Now, lets build the decission tree and develop the confussion matrix. 

```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
confusionMatrix(predictionsA1, myTesting$classe)
```

### Using ML algorithms for prediction: Random Forests

For insample prediction

```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
confusionMatrix(predictionsB1, myTesting$classe)
```

Random Forests yielded better Results, as expected!


#Generating Files to submit as answers for the Assignment:

Finally, using the provided Test Set out-of-sample error.

For Random Forests we use the following formula, which yielded a much better prediction in in-sample:

```{r}
#predictionsB2 <- predict(modFitB1, testing, type = "class")
```

Function to generate files with predictions to submit for assignment

```{r}

#answers = LETTERS[1:20]

#pml_write_files = function(x){
#  n = length(x)
#  for(i in 1:n){
#    filename = paste0("problem_id_",i,".txt")
#    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
#  }
#}

#pml_write_files(predictionsB2)


```
